Data Specialization for a Staged Lambda Calculus

### 2014-10-31 paper/example.tex
rewrite 4.0 - 4.1 with nico

Staging is a technique for using language constructs and types to
explicitly staged a computation according to the frequency or the
arrival time of data.  As an example, suppose that we wish to issue
order statistics queries on a collection of items represented as a
list \texttt{l}.  For this tasks, we can use the expected linear time,
quickselect algorithm~\cite{quickselect}, which takes a
list~\texttt{l} and a rank~\texttt{k} and returns the element with
rank~\texttt{k} in the list.  To this end, the algorithm, whose code
is shown in ~\figref{qs-unstaged} partitions the list into two by
using the first element as a pivot~\footnote{We assume that the list
  is prepermuted to guarantee the expected linear time behavior.} and
recurs down to one of the ``halves'' to find the desired element.  The
``half'' to be considered recursively next is determined by the
relationship of~\texttt{k} to the size of the first half~\texttt{n},
which is returned by the function~\texttt{partition}.

@@

Prior work proposed staging for exactly these kinds of applications
and proposed techniques for writing staged
programs~\cite{staging}. Using techniques essentially the same as
prior work, we can stage our quickselect code by using staged typing
and inserting three keywords~\texttt{next, prev, hold} into the
appropriate places in the code as shown in \figref{qs-staged}. To
write the staged version, the programmer first observes that the
rank~\texttt{k} has now the type $\fut\rmint$ (read ``circle int'')
rather than $\rmint$, indicating that it will be available in the
second stage.  All types that are not of circle type are available in
the first stage.  Naturally the output type, representing the $k$th
largest element of the list, is also $\fut\rmint$, since it is not
computed until the second stage.

>>>>

Staging is a technique for using language constructs and types to
explicitly staged a computation according to the frequency or the
arrival time of data.  As an example, suppose that we wish to issue
order statistics queries on a collection of items represented as a
list \texttt{l}.  For this tasks, we can use the expected linear time,
quickselect algorithm~\cite{quickselect}, which takes a
list~\texttt{l} and a rank~\texttt{k} and returns the element with
rank~\texttt{k} in the list.  To this end, the algorithm, whose code
is shown in ~\ref{fig:qs-unstaged} partitions the list into two by
using the first element as a pivot\footnote{We assume that the list
  is prepermuted to guarantee the expected linear time behavior.} and
recurs down to one of the ``halves'' to find the desired element.  The
``half'' to be considered recursively next is determined by the
relationship of~\texttt{k} to the size of the first half~\texttt{n},
which is returned by the function~\texttt{partition}.

@@

Prior work proposed staging for exactly these kinds of applications
and proposed techniques for writing staged
programs~\cite{staging}. Using techniques essentially the same as
prior work, we can stage our quickselect code by using staged typing
and inserting three keywords~\texttt{next, prev, hold} into the
appropriate places in the code as shown in \ref{fig:qs-staged}. To
write the staged version, the programmer first observes that the
rank~\texttt{k} has now the type $\fut\rmint$ (read ``circle int'')
rather than $\rmint$, indicating that it will be available in the
second stage.  All types that are not of circle type are available in
the first stage.  Naturally the output type, representing the $k$th
largest element of the list, is also $\fut\rmint$, since it is not
computed until the second stage.

### 2014-10-31 paper/splitting.tex
rewrite 4.0 - 4.1 with nico

%\langmono\ has a evaluation judgment $\redsym$, which works in the expected way
%to produce unstaged values.

@@

%As with the dynamics, our stage-splitting algorithm for expressions in \lang\
%takes the form of two mutually-dependent judgments, $\splitonesym$ and
%$\splittwosym$, which respectively split terms that type $@ \bbone$ and $@
%\bbtwo$.

@@

%Since splitting stage \bbtwo\ terms from \lang\ is a simpler operation than
%splitting stage 1 terms, we first describe the mechanics of splitting stage
%\bbtwo, before moving to the more complex case of stage \bbone\ term splitting.

@@

Splitting a stage \bbtwo\ term $\coltwo e A$ produces a \emph{precomputation}
term $p$ and a \emph{resumer} function $l.r$, such that $e$ is equivalent to
$\letin l p r$. This is implemented by the judgment $\splittwosym$ 
(\cref{fig:termSplitTwo}), which operates inductively on the structure of the
term.

For example, the stage \bbtwo\ term
\begin{lstlisting}
2`hold{`1` 1+2 `2`} < hold{`1` 3+4 `2`} - 5`
\end{lstlisting}
splits into a precomputation \verb|(1+2,3+4)| and a resumer
\verb|(L1,L2) => L1 < L2-5|, with the property that the resumer applied to the
precomputation must agree with the original term. This process of hoisting stage
\bbone\ subterms out of $e$ is called \emph{precomputation lifting}, and
intuitively, amounts to computing a data structure which is consumed by $r$.

Since base types and variables have no stage \bbone\ content, their
precomputations are \texttt{()}, and are not used in the resumer. At every other
expression except $\prev$ and $\pause$, splitting bundles the precomputations of the
constituent parts, and then unbundles them with a pattern, with the resumer
precisely resembling the original expression. The $\prev$ and $\pause$ rules depends on
splitting stage \bbone\ terms, which is covered later.

Observe that in the rules for {\tt if}s, {\tt case}s, and functions, the
precomputation is lifted out from within branches. This is the manifestation of
the speculation behavior from the semantics.

%As mentioned before, the only responsibility of $\splittwosym$ is to lift
%out the precomputation from contained stage \bbone\ code into a single
%expression $p$, leaving a resumer $l.r$ to interpret the results of that
%precomputation.

\subsection{(Partial) Value Splitting}

As described above, splitting a stage \bbtwo\ value always produces a trivial
precomputation---either \texttt{()}, or nested tuples thereof. Intuitively,
this is because stage-\bbtwo\ values contain no stage \bbone\ computations.

The same is not true of partial values (by definition, stage \bbone), which \emph{can}
contain stage \bbtwo\ computations, namely, inside $\next$ blocks, which must
contain a variable reference to some future context, and inside lambdas, which
may contain a $\next$ in their body.

Thus, to split partial values, we must separate the contents of $\next$ blocks
from the rest of the term, while maintaining the term's original structure. This
is implemented by the judgment $v \vsplito [v_1,v_2]$ (\ref{fig:valSplit}).
At product, sum, and recursive types, splitting acts recursively, gathering all
subterms' stage $w$ content into $v_w$.
At base types, splitting assigns $v$ entirely to stage \bbone\ (and \texttt{()}
to stage \bbtwo);
conversely, at $\fut$ types, splitting assigns $v$ (which must have the
form $\next~\hat y$) entirely to stage \bbtwo\ (and \texttt{()} to stage
\bbone).
This process of exposing only the relevant structure of a term at each stage is
called \emph{masking}.

Splitting a function is more complex, because function bodies are open terms,
and so are split by the judgments described in the next subsection. Essentially,
functions will split into two functions, whose bodies correspond to the stage
\bbone\ and stage \bbtwo\ components of the original function body.
Moreover, the body of the function may contain a precomputation,
which should be saved and resumer appropriately wherever the function is applied.

%Indeed it would be almost nonsensical to talk of an algorithm for splitting
%stage \bbtwo\ values, since the precondition, that the input is a stage \bbtwo\
%value, is enough to imply the postcondition, that no stage interleaving remains

@@

\subsection{Stage \bbone\ Term Splitting}

@@

Splitting a stage \bbone\ term, in contrast, requires both masking and splitting.
Consider the term 
\begin{lstlisting}
1`(next{`2`hold{`1` 1+2 `2`} < 5`1`},3*4)`
\end{lstlisting}
Masking yields the stage \bbone\ computation \verb|((),3*4)|, while lifting
yields the (stage \bbone) precomputation \verb|(1+2)|. While both computations
must occur at stage \bbone, the former is needed by stage \bbone, while the
latter is needed by stage \bbtwo.

@@

%\subsection {Notable Properties}
%
%\nr{These points will probably get lifted into the preceding sections.}
%
%There are some particularly notable features of this definition of term
%splitting.

%\paragraph {Stage \bbone\ Divergence}
%

%\paragraph {Stage \bbone\ Functions}
%
%The body of a stage \bbone\ function may contain stage \bbtwo\ computation. We
%handle this by splitting the function into two functions: one in the immediate
%result which handles all the stage \bbone\ content of the original, and one in
%the resumer which handles all of the stage \bbtwo\ content. Note that stage
%\bbone\ lambdas are partial values, and so have only a trivial precomputation.
%
%\paragraph {Typing the Boundary}
%
%\TODO add the example which shows that we need dynamic types

@@

\TODO say something about correctness criteria for splitting? (see
\texttt{obsolete/correctness.tex})

@@

\TODO don't use that word

>>>>

@@

@@

@@

Splitting a stage \bbtwo\ term $\coltwo e A$ produces a \emph{precomputation}
term $p$ and a \emph{resumer} function $l.r$, which contain all stage \bbone\
and \bbtwo\ subcomputations of $e$, respectively. Since stage \bbtwo\
computations can depend on stage \bbone, $r$ can depend on $p$ via the variable
$l$. The correctness of splitting thus requires that $e$ is equivalent to
$\letin l p r$.

Splitting, as implemented by the judgment $e \splittwosym [p,l.r]$
(\cref{fig:termSplitTwo}), is guided by the structure of the term $e$. 
For example, the stage \bbtwo\ term
\begin{lstlisting}
2`hold{`1` 1+2 `2`} < (hold{`1` 3+4 `2`} - 5)`
\end{lstlisting}
has two subterms, the latter of which itself has two subterms. Two are $\pause$s
containing stage \bbone\ computations, while the third has no stage \bbone\
content. The precomputation \emph{lifts} these computations in the same tree
structure: \verb|(1+2,(3+4,()))|.%
\footnote{In later examples, we will optimize out unnecessary \texttt{()}s.}
The resumer is a function which computes the stage \bbtwo\ value of the original
term by mapping the results of the precomputation in place of the $\pause$s:
\verb|(L1,(L2,_)) => L1 < (L2 - 5)|.

In general, constants and variables have no stage \bbone\ content, so their
precomputations are \texttt{()}, and their resumers return the original term's
value. Non-staging constructs with $n$ subterms have precomputations which are
$n$-tuples of their subterms' precomputations, and resumers which pass each
component to the resumer of the respective subterm. (This behavior at
\texttt{case} and functions gives rise to speculative execution of
precomputations from underneath binders.) The $\prev$ and $\pause$ rules depends
on splitting stage \bbone\ terms, which is covered later.

\subsection{Partial Value Splitting}

%while maintaining the term's original structure.

Stage \bbtwo\ values can contain no stage \bbone\ subcomputations, but the
reverse is not true of partial values (at stage \bbone). These \emph{can}
contain stage \bbtwo\ computations, both inside $\next$ blocks, which must
contain a variable reference to some future context, and inside lambdas, which
may contain a $\next$ in their body.

Thus, to split partial values, we must separate the contents of $\next$ blocks
from the rest of the term. This is implemented by the judgment $v \vsplito
[v_1,v_2]$ (\ref{fig:valSplit}).

No dependence here.

For example, the stage \bbone\ term
\begin{lstlisting}
1`inl(3, next{`2`yhat`1`})`
\end{lstlisting}
splits into
\verb|inl(3,())| and \texttt{((),$\mathtt{\hat{y}}$)}.

sum types

In general,
at product and recursive types, splitting acts
recursively, gathering all subterms' stage $w$ content into $v_w$. At base
types, splitting assigns $v$ entirely to stage \bbone\ (and \texttt{()} to stage
\bbtwo); conversely, at $\fut$ types, splitting assigns $v$ (which must have the
form $\next~\hat y$) entirely to stage \bbtwo\ (and \texttt{()} to stage
\bbone).

This process of exposing only the relevant structure of a term at each stage is
called \emph{masking}.

Splitting a function is more complex, because function bodies are open terms,
and so are split by the judgments described in the next subsection. Essentially,
functions will split into two functions, whose bodies correspond to the stage
\bbone\ and stage \bbtwo\ components of the original function body.
Because that body may contain a precomputation, the generated function signatures must properly emit and resume with that precomputation.

%As described above, splitting a stage \bbtwo\ value always produces a trivial
%precomputation---either \texttt{()}, or nested tuples thereof.

@@

\subsection{Stage \bbone\ Splitting}

\crem{value splitting figure is missing injections}

\TODO
\begin{enumerate}
\item Examples
\item Lifting (and masking)
\item Correctness? (Does this require partial value splitting?)
\end{enumerate}

\begin{theorem}[Correctness of splitting at $\bbone$]
If $\vdash e:A~@~\bbone$, then (by splitting)
\begin{itemize}
\item $\vdash e : A \splitonesym [c,l.r]$
\item $\reduce {c} {(v_1,u)}$
\item $\reduce {(\letin{l}{u}{r})} v_S$
\end{itemize}
if and only if (by the staged dynamic semantics)
\begin{itemize}
\item $\diaone [] e {\xi;v}$
\item $\dom \xi \vdash v \vsplito [v_1,v_2]$
\item $\reify \xi{v_2}q$
\item $q \mathbin{\tworedsym} v_D$
\end{itemize}
and if so, then $v_D \sim v_S$.
\end{theorem}

@@

Splitting a stage \bbone\ term, in contrast, requires both masking and lifting.
Consider the term 
\begin{lstlisting}
1`(next{`2`hold{`1` 1+2 `2`} < 5`1`},3*4)`
\end{lstlisting}
Masking yields the stage \bbone\ computation \verb|((),3*4)|, while lifting
yields the (stage \bbone) precomputation \verb|(1+2)|. While both computations
must occur at stage \bbone, the former is needed by stage \bbone, while the
latter is needed by stage \bbtwo.

@@

@@

@@

\crem{don't say ``cost''}
